---
title: "DSS Fall 2017 Seminar 4"
output: html_notebook
authors: Chris & Christy
---

## Introduction
This is the notebook for the final `R` seminar session for Fall 2017.

## Dependencies

You will also need a number of other packages that we've worked with in the past sessions:

```{r}
# install using install.packages("tidyverse")
library(broom)         # tidying data
library(dplyr)         # wrangling data
library(ggplot2)       # plotting data

# install individually using install.packages()
library(mapproj)       # tools for working with map projections
library(maptools)      # tools for working with spatial data
library(RColorBrewer)  # color ramps
library(rgdal)         # work with shapefiles
library(tidycensus)    # accessing spatial data

# install using the devtools package
# devtools::install_github("chris-prener/stlData")
library(stlData)       # St. Louis spatial data
```

## Review - Getting Back to Where We Left Off
We use the `rgdal` package to read shapefiles into `R`. These data must be in the **top-level** of your `R` project directory. To import one of our files, the `stlRivers` shapefile, we use the `readOGR()` function. We need to specify `dsn = "."` for each use of the function, and the layer should match the name of your shapefile.

Once imported, these data will appear under the "Values" section of the global environment as an object named `import`. However, they are not in a data frame structure. We need to use the `broom` package's `tidy()` function to dis-aggregate these data and store them in a data frame. We'll do this based on a common ID variable named `GEOID`:

```{r}
import <- readOGR(dsn = ".", layer = "stlTract")
rivers <- tidy(import, region = "GEOID")
```

We also need to load our data on lead poisoning and 


## Plotting Data
### Merging Data
The first step towards plotting our down data is to load some tabular data that we want to map, and quickly explore it using the `str()` function. We are looking for a unique identification number for each row in the tabular data.

```{r}
lead <- stlLead
str(stlLead)
```

Note how `geoID` is a numeric variable. In order to map these data, we have to make two fundamental changes. The first is to rename `geoID` to `id` so that it matches the name of the similar variable in our `stlTract` data frame. The second change is to convert `geoID` from numeric to string. We identified both the name change to `id` and the fact that `id` is a character variable when we explored the newly created `stlTract` data above.

To rename variables, we can use the `dplyr` package's `rename()` function. To re-code variables, we can use the `dplyr` package's `mutate()` function.

To make these changes, we are going to use a pipe (`%>%`). Pipes make our code more readable - they can be read in paragraph or list-like form:

1. We will take our `lead` data, *then*...
2. We will rename the variable `geoID` to `id` in those `lead` data, *then*...
3. We will recede `id` to character, and assign those changes onto the `id` variable (as opposed to making a new variable), *then*...
4. We will assign all of these changes back onto the `lead` data frame (as opposed to making a new data frame object).

Each time we say the word "then", we use a pipe (`%>%`) operator.

```{r}
lead %>%
  rename(id = geoID) %>%
  mutate(id = as.character(id)) -> lead
```

With these changes complete, we want to combine the `stlTract` data with the `lead` data. We'll use the variable that is now named and formatted identically in both data frames - `id`. The `left_join()` function from `dplyr` will take the `stlTract` data and add the `lead` variables to `stlTract`'s right side, assigning the `stlTract` vertices the data from the matching census tract in the `lead` data. We combine these two data frames into a new data frame named `leadMap`:

```{r}
leadMap <- left_join(stlTract, lead, by = "id")
```

Now, you try replicating the three previous code blocks by:

1. loading the `stlInsurance` data stored in the `stlData` package to a data frame named `insurance`
2. renaming `geoID` to `id`
3. re-coding `id` from numeric to string using `mutate()`
4. assigning these changes to the `insurance` data frame
5. using the `left_join()` function to combine the `stlTract` data with the `insurance` data into a new data frame named `insMap`:

```{r}




```

### Plotting Quantities
With our data now joined, we are ready to map them! First, we'll use the `leadMap` data frame to map the percentage of elevated blood lead level tests per census tract. These data are stored in the `pctElevated` variable. We use the `fill` option within the aesthetic mapping to accomplish this. It is the *only* change from our previous map, other than switching the data source.

```{r}
ggplot() +
    geom_polygon(data = leadMap, mapping = aes(x = long, y = lat, group = id, 
                                               fill = pctElevated)) +
    coord_map()
```

Now you trying creating a plot of the `pctUnins` data - the percent uninsured - in the `insMap` data frame. We've started the code here, you just need to add in the `fill = variableName` in the appropriate place:

```{r}
ggplot() +
    geom_polygon(data = insMap, mapping = aes(x = long, y = lat, group = id)) +
    coord_map()
```

### Fixing the Color Ramp
The default color ramps that `R` presents are difficult to interpret. The `RColorBrewer` package you installed earlier provides us with a range of color ramps to select from. You can view these using the following function:

```{r}
display.brewer.all()
```

We'll start by using the `scale_fill_distiller()` function to choose the `"Greens"` palette: 

```{r}
ggplot() +
    geom_polygon(data = leadMap, aes(x = long, y = lat, group = id, fill = pctElevated)) +
    scale_fill_distiller(palette = "Greens") +
    coord_map()
```

One issue with this color ramp above is that the lightest colors correspond to the worst outcomes. Typically, we want to highlight those areas with the highest value (brightest) colors in our ramp. We can create a more intuitive map by adding the `trans = "reverse"` option into the `scale_fill_distiller()` function:

```{r}
ggplot() +
    geom_polygon(data = leadMap, aes(x = long, y = lat, group = id, fill = pctElevated)) +
    scale_fill_distiller(palette = "Greens", trans = "reverse") +
    coord_map()
```

Now, you try adding a `"Reds"` color ramp in reverse order to the code we've provided you below, which maps the `pctUnins` variable from the `insMap` data:

```{r}
ggplot() +
    geom_polygon(data = insMap, aes(x = long, y = lat, group = id, fill = pctUnins)) +
    coord_map()
```

## Putting It All Together

At this point, we can combine both the data layer `geom` from the `leadMap` data with the `rivers` data to produce a serviceable map of lead poisoning rates in the City of St. Louis:

```{r}
ggplot() +
    geom_polygon(data = leadMap, aes(x = long, y = lat, group = id, fill = pctElevated)) +
    geom_polygon(data = rivers, mapping = aes(x = long, y = lat, group = id), fill = "#00009d", color = "#00009d") +
    scale_fill_distiller(palette = "Greens", trans = "reverse") +
    coord_map()
```

Now, you try creating a similar map that combines the `rivers` data with the rates of un-insurance in the City of St. Louis:

```{r}



```

## Accessing Demographic Data

the `stlLead` data frame from `stlData` comes with a select amount of demographic data about the city, but there is a wealth of other information that is available through the U.S. Census Bureau. Obtaining these data used to be an arduous process. However, Kyle Walker, a professor at TCU, has written a package called `tidycensus` to make obtaining these data easier.

### Setting Up tidyensus

Before we use `tidycensus`, we need to take the [API key](https://api.census.gov/data/key_signup.html) that we asked you to obtain and enter it into the area between the double quotes:

```{r}
census_api_key("YOUR API KEY GOES HERE")
```

You should only have to do this once per computer.

### Viewing Possible Variables

There are thousands of possible variables within the Census Bureau's products. We're going to pull a specific variable, the median household income, from the 2015 American Community Survey's (ACS) 5-year estimates for City of St. Louis census tracts. We need to specify the variable we want as well as the Census data product, its vintage, and the geographic parameters we are interested in:

```{r}
medianInc <- get_acs(variables = "B19013_001", survey = "acs5", year = 2015, 
                     geography = "tract", state = "MO", county = 510)
```

`tidycensus` provides a number of tools for looking up different measures in the ACS as well as the decennial census that are worth exploring on your own.

With those data downloaded, we can follow a similar process as we did before for mapping these data. We'll rename `GEOID` to `id`, use the `left_join()` function from `dplyr` to combine the median income data with our geometric data for St. Louis census tracts, and then build a plot around those data layered with the river data. We've selected a different `RColorBrewer` ramp here - `"BuPu"` (Blue and Purple).

```{r}
medianInc <- rename(medianInc, id = GEOID)

incMap <- left_join(stlTract, medianInc, by = "id")

ggplot() +
    geom_polygon(data = incMap, aes(x = long, y = lat, group = id, fill = estimate)) +
    geom_polygon(data = rivers, mapping = aes(x = long, y = lat, group = id), fill = "#00009d", color = "#00009d") +
    scale_fill_distiller(palette = "BuPu", trans = "reverse") +
    coord_map()
```

Now, you try downloading the variable `"B01001B_001"` (number of African Americans) from the 2015 5-year ACS for City of St. Louis census tracts. We've got you started below:

```{r}
countBlack <- get_acs(variables = "", survey = "acs5", year = 2015, 
                     geography = "tract", state = "MO", county = 510)
```

With those data downloaded, follow the same process as we did in the previous example:

1. rename `GEOID` to `id`, 
2. use the `left_join()` function from `dplyr` to combine the median income data with our geometric data for St. Louis census tracts, 
3. and then build a plot around those data layered with the river data;
4. use the `"RdPu"` (Red Purple) color ramp for your data

```{r}



```
