---
title: "DSS Fall 2017 Seminar 4"
output: html_notebook
authors: Chris & Christy
---

## Introduction
This is the notebook for the final `R` seminar session for Fall 2017.

## Dependencies

You will also need a number of other packages that we've worked with in the past sessions:

```{r}
# install using install.packages("tidyverse")
library(broom)         # tidying data
library(dplyr)         # wrangling data
library(ggplot2)       # plotting data

# install individually using install.packages()
library(mapproj)       # tools for working with map projections
library(maptools)      # tools for working with spatial data
library(RColorBrewer)  # color ramps
library(rgdal)         # work with shapefiles
library(tidycensus)    # accessing spatial data

# install using the devtools package
# devtools::install_github("chris-prener/stlData")
library(stlData)       # St. Louis spatial data
```

## Review - Getting Back to Where We Left Off
### Reading in Geometric Data
We use the `rgdal` package to read shapefiles into `R`. These data must be in the **top-level** of your `R` project directory. To import one of our files, the `stlRivers` shapefile, we use the `readOGR()` function. We need to specify `dsn = "."` for each use of the function, and the layer should match the name of your shapefile.

Once imported, these data will appear under the "Values" section of the global environment as an object named `import`. However, they are not in a data frame structure. We need to use the `broom` package's `tidy()` function to dis-aggregate these data and store them in a data frame. We'll do this based on a common ID variable named `GEOID`:

```{r}
import <- readOGR(dsn = ".", layer = "stlTract")
stlTract <- tidy(import, region = "GEOID")
```

### Lead Poisoning Data

We also need to load our data on lead poisoning and prepare it to be joined with the shapefile data:

```{r}
# load data from package into global environment
lead <- stlLead

# prepare for join
lead %>%
  rename(id = geoID) %>%
  mutate(id = as.character(id)) -> lead

# join tabular and geometric data
leadMap <- left_join(stlTract, lead, by = "id")
```

Once the join is complete, we can use `ggplot2` to map our data:

```{r}
ggplot() +
    geom_polygon(data = leadMap, mapping = aes(x = long, y = lat, group = id, fill = pctElevated)) +
    coord_map()
```

### Health Insurance Data

We can repeat the same process with the uninsurance data:

```{r}
# load data from package into global environment
unins <- stlInsurance

# prepare for join
unins %>%
  rename(id = geoID) %>%
  mutate(id = as.character(id)) -> unins

# join tabular and geometric data
insMap <- left_join(stlTract, unins, by = "id")

# make an initial map
ggplot() +
    geom_polygon(data = insMap, mapping = aes(x = long, y = lat, group = id, fill = pctUnins)) +
    coord_map()
```

## Fixing the Color Ramp
The default color ramps that `R` presents are difficult to interpret. The `RColorBrewer` package you installed earlier provides us with a range of color ramps to select from. You can view these using the following function:

```{r}
display.brewer.all()
```

We'll start by using the `scale_fill_distiller()` function to choose the `"Greens"` palette: 

```{r}
ggplot() +
    geom_polygon(data = leadMap, aes(x = long, y = lat, group = id, fill = pctElevated)) +
    scale_fill_distiller(palette = "Greens") +
    coord_map()
```

One issue with this color ramp above is that the lightest colors correspond to the worst outcomes. Typically, we want to highlight those areas with the highest value (brightest) colors in our ramp. We can create a more intuitive map by adding the `trans = "reverse"` option into the `scale_fill_distiller()` function:

```{r}
ggplot() +
    geom_polygon(data = leadMap, aes(x = long, y = lat, group = id, fill = pctElevated)) +
    scale_fill_distiller(palette = "Greens", trans = "reverse") +
    coord_map()
```

Now, you try adding a `"Reds"` color ramp in reverse order to the code we've provided you below, which maps the `pctUnins` variable from the `insMap` data:

```{r}
ggplot() +
    geom_polygon(data = insMap, aes(x = long, y = lat, group = id, fill = pctUnins)) +
    coord_map()
```

## Removing Some of the ggplot2 Elements

To get the plot ready for presentation, we want to add a couple of details and remove some of the `ggplot` elements. The details we want to add are:

* a title,
* a subtitle,
* and a clearer title for the legend. 

We can add all of these using the `labs()` function, which accepts arguments for (among other things) `title`, `subtitle`, and `fill`. Note the commas after each argument, and the use of `\n` to add a line break in the title. We've also changed the color ramp to the `"Purples"` palette from `RColorBrewer`:

```{r}
ggplot() +
    geom_polygon(data = leadMap, aes(x = long, y = lat, group = id, fill = pctElevated)) +
    scale_fill_distiller(palette = "Purples", trans = "reverse") +
    coord_map() +
    labs(
      title = "Percentage of Children Tested \nwith Blood Lead Levels >5µg/dL", 
      subtitle = "St. Louis, Missouri",
      fill = "% Elevated")
```

The gray background and the latitude and longitude information is not particularly informative here. We can remove it by adding the `theme_void()` function to our `ggplot()` call:

```{r}
ggplot() +
    geom_polygon(data = leadMap, aes(x = long, y = lat, group = id, fill = pctElevated)) +
    scale_fill_distiller(palette = "Purples", trans = "reverse") +
    coord_map() +
    labs(
      title = "Percentage of Children Tested \nwith Blood Lead Levels >5µg/dL", 
      subtitle = "St. Louis, Missouri",
      fill = "% Elevated") +
    theme_void()
```

Try combining both of these alterations to the insurance data. We've given you some code to get started. The changes you should make:

* Add a title describing what the map shows
* Add a subtitle noting that the data are for St. Louis, Missouri
* Add a better legend title
* Remove the `ggplot2` theme elements

```{r}
ggplot() +
    geom_polygon(data = insMap, aes(x = long, y = lat, group = id, fill = pctUnins)) +
    scale_fill_distiller(palette = "Oranges", trans = "reverse") +
    coord_map()
```

## Accessing Demographic Data

the `stlLead` data frame from `stlData` comes with a select amount of demographic data about the city, but there is a wealth of other information that is available through the U.S. Census Bureau. Obtaining these data used to be an arduous process. However, Kyle Walker, a professor at TCU, has written a package called `tidycensus` to make obtaining these data easier.

### Setting Up tidyensus

Before we use `tidycensus`, we need to take the [API key](https://api.census.gov/data/key_signup.html) that we asked you to obtain and enter it into the area between the double quotes:

```{r}
census_api_key("YOUR API KEY GOES HERE")
```

You should only have to do this once per computer.

### Downloading data

There are thousands of possible variables within the Census Bureau's products. We're going to pull a specific variable, the median household income, from the 2015 American Community Survey's (ACS) 5-year estimates for City of St. Louis census tracts. We need to specify the variable we want as well as the Census data product, its vintage, and the geographic parameters we are interested in:

```{r}
medianInc <- get_acs(variables = "B19013_001", survey = "acs5", year = 2015, 
                     geography = "tract", state = "MO", county = 510)
```

`tidycensus` provides a number of tools for looking up different measures in the ACS as well as the decennial census that are worth exploring on your own.

With those data downloaded, we can follow a similar process as we did before for mapping these data. We'll rename `GEOID` to `id`, use the `left_join()` function from `dplyr` to combine the median income data with our geometric data for St. Louis census tracts, and then build a plot around those data layered with the river data. We've selected a different `RColorBrewer` ramp here - `"BuPu"` (Blue and Purple).

```{r}
medianInc <- rename(medianInc, id = GEOID)

incMap <- left_join(stlTract, medianInc, by = "id")

ggplot() +
    geom_polygon(data = incMap, aes(x = long, y = lat, group = id, fill = estimate)) +
    scale_fill_distiller(palette = "BuPu", trans = "reverse") +
    coord_map() +
    labs(
      title = "Median Income by Census Tract, 2011-2015", 
      subtitle = "St. Louis, Missouri",
      fill = "Median \nIncome") +
    theme_void()
```

Now, you try downloading the variable `"B01001B_001"` (number of African Americans) from the 2015 5-year ACS for City of St. Louis census tracts. We've got you started below:

```{r}
countBlack <- get_acs(variables = "", survey = "acs5", year = 2015, 
                     geography = "tract", state = "MO", county = 510)
```

With those data downloaded, follow the same process as we did in the previous example:

1. rename `GEOID` to `id`, 
2. use the `left_join()` function from `dplyr` to combine the median income data with our geometric data for St. Louis census tracts, 
3. and then build a plot using the `"RdPu"` (Red Purple) color ramp for your data

```{r}



```
